---
title: "scMET scalability on running windows"
author: "C.A.Kapourani & R. Argelaguet"
output: 
  html_notebook:
    df_print: paged
    highlight: haddock
    number_sections: yes
    theme: cerulean
    toc: yes
---

# Load libraries and settings
```{r}
suppressPackageStartupMessages(library(scMET))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ggrepel))
suppressPackageStartupMessages(library(ggpubr))

## I/O
source("../load_settings.R")
out_dir <- "~/datasets/scMET_ms/ecker2017/scalability/"
if (!dir.exists(out_dir)) { dir.create(out_dir, recursive = TRUE) }
```

# Running times
Taken from logs of scMET. Should automate this, for now adding running times, manually.
```{r}
# In hours
size_1k_all_cells <- c(30, 33, 23, 30, 31, 29, 32, 27, 28, 24, 33, 24, 28, 21, 18, 19, 23, 23, 13, 8)
size_2k_all_cells <- c(31, 32, 24, 25, 27, 32, 32, 28, 24, 27, 31, 20, 30, 21, 19, 18, 16, 21, 12, 10)
```


# Load methylation data
```{r}
####################################
## Load / filter methylation data ##
Y <- list()
for (i in anno_id){
Y[[i]] <- read_filter_ecker_data(filename = sprintf("%s/%s.tsv.gz", io$data_parsed, i),
                                 opts = opts, is_differential = FALSE)
}
feat_names <- lapply(Y, function(x) unique(x$Feature))
```

## Read and keep annotation feature information
```{r}
anno_dt <- list()
for (i in 1:length(anno_id)) {
  anno_dt[[i]] <- fread(paste0(anno_id[i], ".bed.gz")) %>%
    .[, c(2, 3, 5)] %>% setnames(c("start", "end", "id")) %>%
    .[id %in% feat_names[[anno_id[i]]]] %>% .[, length := (end - start)] %>%
    .[, anno_name := anno_name[i]] %>%
    .[, anno := anno_id[i]]
}
anno_dt <- rbindlist(anno_dt)
```

## Plot annotation feature length distribution
```{r}
pdf(sprintf("%s/feature_lengths.pdf", out_dir), width = 7, height = 4)
ggboxplot(anno_dt, x =  "anno_name", y = "length", ylab = "Feature length", 
          xlab = "", fill = "cornflowerblue") +
  coord_cartesian(ylim = c(0, 2500))
dev.off()
```


# Relationship of feature length with HVF/ mean methylation and overdispersion
```{r}
hvfdir <- "~/datasets/scMET_ms/ecker2017/all_cells/hvf/hits/"
hvf_dt <- anno_id %>%
  map(~ fread(sprintf("%s/hvf_%s_epsilon.txt.gz", hvfdir, .))) %>%
  rbindlist
```

## Merge data
```{r}
hvf_merged <- merge(hvf_dt, anno_dt, by = c("anno", "id"))
```


## Plots
```{r, fig.height=3, fig.width=3}
set.seed(1234)
gg_list <- list()
for (an in anno_id) {
  title <- ifelse(an == "distal_H3K27ac_cortex", "Distal H3K27ac", "H3K4me1") 
  #df <- df_merged[anno == an] %>% .[sample(x = NROW(df_merged), size = 8000), ] 
  df <- hvf_merged[anno == an]
  gg1 <- ggplot(df, aes(x = log(length), y = mu)) +
    geom_hex(bins = 70) +
    #geom_point(size = 1, shape = 21, stroke = 0.2, fill = "#E69F00") + 
    xlab("Log feature length") + ylab("Mean methylation") +
    ggtitle(title) +
    viridis::scale_fill_viridis() +
    viridis::scale_color_viridis() +
    #coord_cartesian(xlim = c(0, 2500)) +
    theme_classic() 
  gg2 <- ggplot(df, aes(x = log(length), y = epsilon)) +
    geom_hex(bins = 50) +
    #geom_point(size = 1, shape = 21, stroke = 0.2, fill = "#E69F00") + 
    xlab("Log feature length") + ylab("Residual Overdispersion") + 
    viridis::scale_fill_viridis() +
    viridis::scale_color_viridis() +
    coord_cartesian(ylim = c(-2.8, 2.5)) +
    theme_classic()
  
  gg3 <- ggplot(df, aes(x = log(length), y = tail_prob)) +
    geom_hex(bins = 50) +
    #geom_point(size = 1, shape = 21, stroke = 0.2, fill = "#E69F00") + 
    xlab("Log feature length") + ylab("HVF posterior tail probability") + 
    viridis::scale_fill_viridis() +
    viridis::scale_color_viridis() +
    theme_classic()
  
  gg_list[[an]] <- cowplot::plot_grid(gg1, gg2, gg3, ncol = 3)
}

library(patchwork)
tmp <- (gg_list[[1]]/gg_list[[2]]) + plot_annotation(tag_levels = 'a')
pdf(sprintf("%s/feature_lengths_vs_estimates.pdf", out_dir), width = 14, height = 7)
print(tmp)
dev.off()
```



# Relationship of feature length with DE analysis
```{r}
diffdir <- "~/datasets/scMET_ms/ecker2017/diff/hits/"
diff_dt <- list() 
for (an in anno_id) {
  diff_dt[[an]] <- readRDS(sprintf("%s/%s_vb.rds", diffdir, an))
}
```

## Merge data
```{r}
for (an in anno_id) {
  diff_dt[[an]]$diff_mu_summary <- merge(diff_dt[[an]]$diff_mu_summary, 
                                         anno_dt, by.x = c("feature_name"), by.y = "id")
  diff_dt[[an]]$diff_epsilon_summary <- merge(diff_dt[[an]]$diff_epsilon_summary, 
                                         anno_dt, by.x = c("feature_name"), by.y = "id")
  diff_dt[[an]]$diff_gamma_summary <- merge(diff_dt[[an]]$diff_gamma_summary, 
                                         anno_dt, by.x = c("feature_name"), by.y = "id")
}
```


## Plots
```{r, fig.height=3, fig.width=3}
set.seed(1234)
gg_list <- list()
for (an in anno_id) {
  title <- ifelse(an == "distal_H3K27ac_cortex", "Distal H3K27ac", "H3K4me1") 
  gg1 <- ggplot(as.data.frame(diff_dt[[an]]$diff_mu_summary), 
                aes(x = log(length.y), y = mu_LOR)) +
    geom_hex(bins = 70) +
    #geom_point(size = 1, shape = 21, stroke = 0.2, fill = "#E69F00") + 
    xlab("Log feature length") + ylab("LOR (mean methylation)") +
    ggtitle(title) +
    viridis::scale_fill_viridis() +
    viridis::scale_color_viridis() +
    #coord_cartesian(xlim = c(0, 2500)) +
    theme_classic() 
  gg2 <- ggplot(as.data.frame(diff_dt[[an]]$diff_epsilon_summary), 
                aes(x = log(length), y = epsilon_change)) +
    geom_hex(bins = 50) +
    #geom_point(size = 1, shape = 21, stroke = 0.2, fill = "#E69F00") + 
    xlab("Log feature length") + ylab("Change in residual overdispersion") + 
    viridis::scale_fill_viridis() +
    viridis::scale_color_viridis() +
    coord_cartesian(ylim = c(-2.8, 2.5)) +
    theme_classic()
  
  gg_list[[an]] <- cowplot::plot_grid(gg1, gg2, ncol = 2)
}

library(patchwork)
tmp <- (gg_list[[1]]/gg_list[[2]]) + plot_annotation(tag_levels = 'a')
pdf(sprintf("%s/feature_lengths_vs_diff.pdf", out_dir), width = 9, height = 7)
print(tmp)
dev.off()
```



# Relationship of feature length and CpGs with normalised dispersion estimates
```{r}
datadir <- "~/datasets/scMET_ms/ecker2017/all_cells/data/"
disp_dt <- anno_id %>%
  map(~ fread(sprintf("%s/%s_norm_dispersion_beta.txt.gz", datadir, .))) %>%
  rbindlist %>%
  setnames(c("Feature"), c("id"))
```

## Merge data
```{r}
disp_merged <- merge(disp_dt, anno_dt, by = c("anno", "id"))
```


## Plots
```{r, fig.height=3, fig.width=3}
set.seed(1234)
gg_list <- list()
for (an in anno_id) {
  title <- ifelse(an == "distal_H3K27ac_cortex", "Distal H3K27ac", "H3K4me1") 
  #df <- df_merged[anno == an] %>% .[sample(x = NROW(df_merged), size = 8000), ] 
  df <- disp_merged[anno == an]
  gg1 <- ggplot(df, aes(x = log(length), y = dispersion_norm_sign)) +
    geom_hex(bins = 70) +
    #geom_point(size = 1, shape = 21, stroke = 0.2, fill = "#E69F00") + 
    xlab("Log feature length") + ylab("Normalised dispersion") +
    ggtitle(title) +
    viridis::scale_fill_viridis() +
    viridis::scale_color_viridis() +
    #coord_cartesian(xlim = c(0, 2500)) +
    theme_classic() 
  gg2 <- ggplot(df, aes(x = log(length), y = cv)) +
    geom_hex(bins = 50) +
    #geom_point(size = 1, shape = 21, stroke = 0.2, fill = "#E69F00") + 
    xlab("Log feature length") + ylab("Coefficient of variation") + 
    viridis::scale_fill_viridis() +
    viridis::scale_color_viridis() +
    #coord_cartesian(ylim = c(-2.8, 2.5)) +
    theme_classic()
  
  gg_list[[an]] <- cowplot::plot_grid(gg1, gg2, ncol = 2)
}

library(patchwork)
tmp <- (gg_list[[1]]/gg_list[[2]]) + plot_annotation(tag_levels = 'a')
pdf(sprintf("%s/norm_disp_feature_lengths_vs_estimates.pdf", out_dir), width = 9, height = 7)
print(tmp)
dev.off()
```

## Correlation with CpG coverage
```{r}
set.seed(1234)
gg_list <- list()
for (an in anno_id) {
  title <- ifelse(an == "distal_H3K27ac_cortex", "Distal H3K27ac", "H3K4me1") 
  #df <- df_merged[anno == an] %>% .[sample(x = NROW(df_merged), size = 8000), ] 
  df <- Y[[an]]
  df <- df %>% .[, .(cpgs = mean(total_reads)), by = .(Feature)] %>%
    setnames(c("id", "cpgs"))
  df <- merge(df, disp_merged, by = c("id"))
  
  gg1 <- ggplot(df, aes(x = log(cpgs), y = dispersion_norm_sign)) +
    geom_hex(bins = 70) +
    #geom_point(size = 1, shape = 21, stroke = 0.2, fill = "#E69F00") + 
    xlab("Number of CpGs") + ylab("Normalised dispersion") +
    ggtitle(title) +
    viridis::scale_fill_viridis() +
    viridis::scale_color_viridis() +
    #coord_cartesian(xlim = c(0, 2500)) +
    theme_classic() 
  gg_list[[an]] <- gg1
}

library(patchwork)
tmp <- (gg_list[[1]] | gg_list[[2]]) + plot_annotation(tag_levels = 'a')
pdf(sprintf("%s/norm_disp_cpgs_vs_estimates.pdf", out_dir), width = 9, height = 4)
print(tmp)
dev.off()
```


