---
title: "Window based HVF calling on residual overdispersion"
author: "C.A.Kapourani & R. Argelaguet"
output: 
  html_notebook:
    df_print: paged
    highlight: haddock
    number_sections: yes
    theme: cerulean
    toc: yes
---

# Load libraries and settings
```{r}
suppressPackageStartupMessages(library(scMET))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ggrepel))
suppressPackageStartupMessages(library(ggpubr))

## I/O
source("../../load_settings.R")
data_dir <- "~/datasets/scMET_ms/ecker2017/all_cells/data/"
marker_genes_file <- "~/datasets/scMET_ms/ecker2017/metadata/ecker2017_marker_genes.csv"
out_dir <- "~/datasets/scMET_ms/ecker2017/all_cells/hvf_window/"
hits_dir <- "~/datasets/scMET_ms/ecker2017/all_cells/hvf_window/hits/"
if (!dir.exists(out_dir)) { dir.create(out_dir, recursive = TRUE) }
if (!dir.exists(hits_dir)) { dir.create(hits_dir, recursive = FALSE) }

## Options
# Perform HVF calling eithr using residual overdispersion (epsilon)
# or overdispersion (gamma)
mode <- "epsilon"

chr <- paste0("chr", c(seq(1, 19), "X"))
#chr <- "chr19"
# Define genomic contexts
annot <- list(
  c("window10000_step10000" = "Window 1kb"),
  c("window20000_step20000" = "Window 2kb")
)
```

<!-- # Read per chromosome and combine results -->
<!-- ```{r} -->
<!-- for (i in 1:length(annot)) { -->
<!--   df <- list() -->
<!--   for (ch in chr) { -->
<!--     df[[ch]] <- fread(sprintf("%s/%s_%s_vb.txt.gz", data_dir, names(annot[[i]]), ch)) -->
<!--   } -->
<!--   df <- rbindlist(df) -->
<!--   fwrite(df, file = sprintf("%s/%s_vb.txt.gz", data_dir, names(annot[[i]])),  -->
<!--          row.names = FALSE, sep = "\t") -->
<!-- } -->
<!-- ``` -->


# HVF analysis
```{r}
hvf_res <- list()
for (i in 1:length(annot)) {
  obj <- readRDS(sprintf("%s/%s_%s_vb.rds", data_dir, names(annot[[i]]), chr))
  if (mode == "epsilon") {
    hvf_res[[annot[[i]]]] <- scmet_hvf(scmet_obj = obj, delta_e = 0.9, delta_g = NULL,
                                       evidence_thresh = 0.8, efdr = 0.1)
  } else {
    hvf_res[[annot[[i]]]] <- scmet_hvf(scmet_obj = obj, delta_e = NULL, delta_g = 0.5,
                                       evidence_thresh = 0.8, efdr = 0.1)
  }
}
rm(obj)
```

# Summary of HVF analysis
Obtain summary of HVF analysis and also store results locally
```{r}
hvf_summary <- list()
for (i in 1:length(annot)) {
  hvf_summary[[annot[[i]]]] <- hvf_res[[annot[[i]]]]$hvf$summary %>% as.data.table %>%
    setnames("feature_name","id") %>%
    .[, anno := names(annot[[i]])]
  if (mode == "epsilon") {
    hvf_summary[[annot[[i]]]] <- hvf_summary[[annot[[i]]]] %>% setorder(-tail_prob, -epsilon)
  } else {
    hvf_summary[[annot[[i]]]] <- hvf_summary[[annot[[i]]]] %>% setorder(-tail_prob, -gamma)
  }
  fwrite(hvf_summary[[i]], 
         file = sprintf("%s/hvf_%s_%s_%s.txt.gz", hits_dir, names(annot[[i]]), chr, mode))
  hvf_summary[[annot[[i]]]] %>% .[, anno_name := annot[[i]] ]
}
```


# Mean overdispersion relationship 
Below we show the mean-overdispersion relationship together with the nonlinear trend fitted by scMET, from which we can extract the residual overdispersion estimates.
```{r, echo=FALSE, results='hide', message=FALSE, fig.width=5, fig.height=2}
set.seed(123)
for (an in 1:length(annot)) {
  gg1 <- scmet_plot_mean_var(obj = hvf_res[[annot[[an]]]], y = "gamma", 
                             task = NULL, show_fit = TRUE, nfeatures = 4500)
  gg2 <- scmet_plot_mean_var(obj = hvf_res[[annot[[an]]]], y = "epsilon", 
                             task = NULL, show_fit = TRUE, nfeatures = 4500)
  pdf(paste0(out_dir, "ecker_meanvar_", names(annot[[an]]), "_", chr, ".pdf"),  
      width = 8.5, height = 3, useDingbats = FALSE)
  plot(cowplot::plot_grid(gg1, gg2, ncol = 2))
  dev.off()
}
```

# Mean - overdispersion plots coloured by HVF 
Below we plot the mean - overdispersion relationship for each genomic context and we colour features as being highly variable (red) or other (grey).
```{r, fig.width=3.7, fig.height=2.8}
set.seed(123)
for (an in 1:length(annot)) {
  ##############
  ## Add labels
  ##############
  hv <- copy(hvf_res[[annot[[an]]]]$hvf$summary) 
  hv <- hv %>% as.data.table %>% setnames(c("feature_name"), c("id")) %>%
    .[, anno := names(annot[[an]])]
  # Map features to nearest genes
  hv <- hv %>% merge(feat2genes[,c("id","anno","gene")], by = c("id","anno"))
  # Store final labels
  lab <- hv[is_variable == TRUE & gene %in% marker_genes$gene_name] %>%
      .[,.SD[which.max(tail_prob)], by = "gene"] %>%
    setorder(-epsilon) %>%
    head(n = 15)
  
  # Total number of hits
  sig_hits <-  hv[is_variable == TRUE, .N]
  nonsig_hits <-  hv[is_variable == FALSE, .N]
  ylim <- max(hv$gamma, na.rm = TRUE) + 0.1
  gg <- scmet_plot_mean_var(obj = hvf_res[[annot[[an]]]], y = "gamma", 
                            task = "hvf", nfeatures = 2500) +
    theme(legend.position = c(0.9, 0.92)) +
    annotate("text", x = 0.17, y = ylim, size = 4, 
             label = sprintf("N=%d (%.1f%%)", 
                             sig_hits, 100*(sig_hits/nonsig_hits))) +
    geom_text_repel(force = 15, data = lab, aes_string(label = "gene"),
                    size = 4, col = "black", segment.color = "black", 
                    segment.size = 0.3, segment.alpha = 0.35, 
                    box.padding = unit(0.5,"lines"), show.legend = FALSE
  )
  pdf(sprintf("%s/ecker_hvf_%s_%s_%s.pdf", out_dir, names(annot[[an]]), chr,
              mode), width = 6, height = 4.6)
  print(gg)
  dev.off()
}
```


# EFDR grid search plots
```{r, echo=FALSE, results='hide', message=FALSE, fig.width=7, fig.height=2.2}
p <- annot %>% 
  map(~ scmet_plot_efdr_efnr_grid(obj = hvf_res[[.]], task = "hvf") + 
        theme(legend.position = "top") + ggtitle(.) ) %>%
  cowplot::plot_grid(plotlist = ., ncol = 2)
pdf(paste0(out_dir, "ecker_efdr_", mode, "_", chr, ".pdf"), width = 9, height = 3.5)
print(p)
dev.off()
```

## Mean methylation versus HVF tail probability
Below we show plots of the relationship between mean methylation and HVF tail probability, that shows the evidence of a feature being called as HVF. In general we do not observe any association, which implies that mean methylation is not a confounder for our HVF calling strategy.
```{r, fig.width=7, fig.height=2.1}
set.seed(123)
p <- annot %>% 
  map(~ scmet_plot_vf_tail_prob(obj = hvf_res[[.]], x = "mu", 
                                task = "hvf", nfeatures = 2000) + 
        ggtitle(.) + theme(legend.position = "top")) %>%
  cowplot::plot_grid(plotlist = ., ncol = 2)
pdf(paste0(out_dir,"ecker_tailprob_vs_mu_", mode, "_", chr, ".pdf"), 
    width = 8, height = 3, useDingbats = FALSE)
print(p)
dev.off()
```


## Variability distribution across all genomic contexts

### Residual overdispersion $\epsilon$
```{r}
to_plot <- rbindlist(hvf_summary)
p <- ggdensity(to_plot, x = "epsilon", y = "..density..", fill = "anno_name", alpha = 0.3) +
  scale_fill_brewer(palette = "Set1") +
  labs(x = expression(paste("Residual overdispersion ", epsilon)), y = "Density") +
  theme(legend.title = element_blank()) +
  xlim(c(-2.7, 2.7))
pdf(sprintf("%s/ecker_density_epsilon_%s.pdf", out_dir, chr), width = 5, 
    height = 3, useDingbats = FALSE)
print(p)
dev.off()
```

### Overdispersion $\gamma$
```{r}
p <- ggdensity(to_plot, x = "gamma", y = "..density..", fill = "anno_name", alpha = 0.3) +
  scale_fill_brewer(palette = "Set1") +
  labs(x = expression(paste("Overdispersion ", gamma)), y = "Density") +
  geom_vline(xintercept = 0.5, color = "black", linetype = "dashed", size = 2) +
  theme(legend.title = element_blank())
pdf(sprintf("%s/ecker_density_gamma_%s.pdf", out_dir, chr), width = 5, 
    height = 3, useDingbats = FALSE)
print(p)
dev.off()
```


### Mean methylation rate $\mu$
```{r}
p <- ggdensity(to_plot, x = "mu", y = "..density..", fill = "anno_name", alpha = 0.3) +
  scale_fill_brewer(palette = "Set1") +
  labs(x = expression(paste("Mean methylation ", mu)), y = "Density") +
  theme( legend.title = element_blank())
pdf(sprintf("%s/ecker_density_mu_%s.pdf", out_dir, chr), width = 5, 
    height = 3, useDingbats = FALSE)
print(p)
dev.off()
```

